---
title: "Ebola outbreak in southwest Uganda: state of the epidemic"
author: "Elisha Are, Jeremy Bingham"
date: "10 January 2020"
output: html_document
theme: sandstone
highlight: tango
bibliography: Challenge_assignment_report.bib
fig_width: 1 
fig_height: 4 
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)

# The following code may be used to install the required packages
# install.packages("devtools")18
# devtools::install_github("reconhub/linelist")
# devtools::install_github("reconhub/reportfactory")
# devtools::install_github("reconhub/epicontacts")
# install.packages("R0")
# *set required packages to automatically install if necessary

library(tcltk)
library(devtools)
library(linelist)
library(tidyverse)
library(lubridate)
library(incidence)
library(R0)


chooseDirectory = function(caption = 'Select data directory') {   #https://stackoverflow.com/questions/48218491/os-independent-way-to-select-directory-interactively-in-r
  if (exists('utils::choose.dir')) {
    choose.dir(caption = caption) 
  } else {
    tk_choose.dir(caption = caption)
  }
}

calculateCarefulCaseFatality <- function(dataset, thresholdTimeToDeath, rounded=TRUE, asPercentage= TRUE){
  lastReportDate = last(dataset$reportDate)
  casesConsidered <- dataset %>%
    subset(as.numeric(difftime(lastReportDate,onsetDate)) >= thresholdTimeToDeath)
  nCases = nrow(casesConsidered)
  nDeaths = nCases - sum(is.na(casesConsidered$deathDate))
  caseFatalityRate = nDeaths/nCases
 
  #* if (nrow(casesConsidered==0)){
  #   print("There are no cases in this dataset which are eligible.")
  #   return(NA)
  # }
  if(rounded){
    if(asPercentage){  
      return(signif(100*caseFatalityRate,3))}
    else{
      return(signif(100*caseFatalityRate,3)/100)
    }
  }
  if(asPercentage){
    return(100*caseFatalityRate)}
  return(caseFatalityRate)
}


calculateCrudeCaseFatality <- function(dataset, rounded=TRUE, asPercentage=TRUE){
  nCases = nrow(dataset)
  nDeaths = nCases - sum(is.na(dataset$deathDate))
  caseFatalityRate = nDeaths/nCases
  
  if(rounded){
    if(asPercentage){
      return(signif(100*caseFatalityRate,3))}
    else{
      return(signif(100*caseFatalityRate,3)/100)
    }
  }
  if(asPercentage){
    return(100*caseFatalityRate)}
  return(caseFatalityRate)
}

```

```{r Data import, echo=FALSE}

# User instruction
# By default the program looks for the directory C:/Users/your_username/Desktop/Smarties/Elisha_Jeremy_Challenge_assignment
# If this directory is not found then the program will prompt you to select the directory where the script is stored
# Feel free to permanently set your own directory by changing "wdstring" below

wdstring = paste("C:/Users/",Sys.info()['login'],"/Desktop/Smarties/Elisha_Jeremy_Challenge_Assignment",sep="")

if(dir.exists(wdstring))
{setwd(wdstring)
  }else{chooseDirectory()}


#Import data
if("ebola_2_update.csv" %in% list.files()){ebolaData <- read.csv(file = "ebola_2_update.csv") #importing raw data
}else{
  print("Select the .csv file where the data is stored.")
  chosenFile <- file.choose()
  ebolaData <- read.csv(chosenFile)
}


ebolaDataDateFormatted <- clean_dates(ebolaData)
# ebola_linelist_data_cleann <-  clean_dates(ebola_linelist_data)


transformedData <- ebolaDataDateFormatted %>%
                                    arrange(onsetDate) %>%
                                    mutate (reportingDelay = difftime(reportDate, onsetDate))

transformedData <- transformedData %>%
                                    mutate(timeBeforeDeath = difftime(deathDate,onsetDate))


ebolaCasesConfirmed <- transformedData %>%
  filter(status=='confirmed')

ebolaCasesSuspected <- transformedData %>%
  filter(status=='suspected')

ebolaCasesProbable <- transformedData %>%
  filter(status=='probable')

ebolaCasesInMales <- transformedData %>%
  filter(sex=='male')

ebolaCasesInFemales <- transformedData %>%
  filter(sex=='female')
```


```{r calculate key stats, echo=FALSE}


lengthOfOutbreak = abs(difftime(first(transformedData$onsetDate),last(transformedData$onsetDate)))
lastReportDate = last(transformedData$reportDate)
firstOnsetDate = first(transformedData$onsetDate)
lastOnsetDate = last(transformedData$onsetDate)

conservativeThreshold <- max(as.numeric(transformedData$timeBeforeDeath),na.rm = TRUE)
liberalThreshold <- mean(as.numeric(transformedData$timeBeforeDeath),na.rm = TRUE)+2*sd(as.numeric(transformedData$timeBeforeDeath),na.rm = TRUE)

caseFatalityEstimate = calculateCarefulCaseFatality(transformedData,liberalThreshold)
caseFatalityEstimateUsingMaxTimeToDeath = calculateCarefulCaseFatality(transformedData,conservativeThreshold)
crudeCaseFatality = calculateCrudeCaseFatality(transformedData)
caseFatalityM = calculateCarefulCaseFatality(ebolaCasesInMales, liberalThreshold)
caseFatalityF = calculateCarefulCaseFatality(ebolaCasesInFemales, liberalThreshold)

meanReportingDelay = mean(as.numeric(difftime(transformedData$reportDate,transformedData$onsetDate)))


incidenceToDate = incidence(transformedData$onsetDate)
maxDailyIncidence = max(incidenceToDate$counts)
dateOfMaxIncidence = incidenceToDate$dates[which.max(incidenceToDate$counts)]

weeklyIncidenceToDate = incidence(transformedData$onsetDate,interval = 7)
maxWeeklyIncidence = max(weeklyIncidenceToDate$counts)
weekOfMaxIncidence = which.max(weeklyIncidenceToDate$counts)


```


## Introduction
This report summarises available data from the ongoing Ebola outbreak taking place in Southwest Uganda, ongoing since `r firstOnsetDate`. The most recent case report included in the available data was made on `r lastReportDate`. The mean reporting delay in the data is `r  signif(meanReportingDelay,1)` days. A total of `r nrow(transformedData)` cases (`r nrow(ebolaCasesInMales)` male, `r nrow(ebolaCasesInFemales)` female), (`r nrow(ebolaCasesSuspected)` suspected, `r nrow(ebolaCasesProbable)` probable, `r nrow(ebolaCasesConfirmed)` confirmed) have been reported to date. The estimated case fatality rate is `r caseFatalityEstimate`\%.

## Incidence
The incidence rate is estimated from the date of onset of the disease. The first incidence of Ebola is taken as the reported case with the earliest date of onset. Figure 1 presents the daily incidence of the disease, starting from  `r firstOnsetDate`  to `r lastOnsetDate`. So far, the largest daily incidence of Ebola was recorded on `r dateOfMaxIncidence`, with `r maxDailyIncidence` cases. Incidence values and plots are calculated and generated using the R package "incidence" [@kamvar2019epidemic].

<!-- # ```{r,fig.height = 3, fig.width = 7, echo=FALSE} -->
<!-- # # par(mfrow=c(1,2)) -->
<!-- # # should we count dates on x axis from outbreak onset or in calendar date -->
<!-- # # ** need titles, x label -->
<!-- # # * make incidence plots pretty -->
<!-- # # plot.new() -->
<!-- # plot(incidence(transformedData$onsetDate,interval = 1))  + labs(y = "Daily incidence", x ="Date" ) -->
<!-- ``` -->

Figure 2 below shows the weekly number of reported Ebola cases from the beginning of the epidemic up until the onset week of the last reported case. Week `r weekOfMaxIncidence` of the outbreak witnessed the highest number of Ebola cases so far, with `r maxWeeklyIncidence` cases occurring. 


```{r,fig.height = 3, fig.width = 7, echo = FALSE}

plot(incidence(transformedData$onsetDate,interval = 7))  + labs(y = "Weekly incidence", x = "Date")

```


Figure 3 below shows the cumulative number of Ebola cases from the date of the first reported case to the date the last case was reported.


```{r,fig.height = 3, fig.width = 7, echo = FALSE}
plot(cumulate(incidence(transformedData$onsetDate,interval = 1)))

```
The daily cumulative incidence plot provides a visualisation of the total burden of the epidemic. **

## Case fatality rate

* The estimated case fatality rate is `r caseFatalityEstimate`\%. The case fatality rate among females is `r caseFatalityF`\% while the case fatality rate among males is `r caseFatalityM`\%.

```{r, echo=FALSE}
casesWhichWouldHaveDiedIfTheyWereGoingToDie <- transformedData %>%
                                              subset(as.numeric(difftime(lastReportDate,onsetDate)) >= liberalThreshold)

mutbd = signif(mean(as.numeric(transformedData$timeBeforeDeath),na.rm = TRUE),1)
sdtbd = signif(2*sd(as.numeric(transformedData$timeBeforeDeath),na.rm = TRUE),1)

```


##### Details regarding case fatality estimation:

The case fatality rate is estimated under the assumption that cases for which the date of death is not reported have recovered. As such, recently-reported cases should not be included in this calculation, since they are relatively likely to still be alive regardless of their prospects for recovery, and their inclusion would result in under-estimation of the case fatality.

The case fatality rate is estimated as follows:
The delays between onset and death are calculated for all cases with a date of death. The mean $\mu$ and standard deviation $\sigma$ of these times (in days) is calculated. The most recent cases considered for inclusion in the case fatality must be have occurred at least $\mu+2 \sigma =$ `r mutbd` $+$ `r sdtbd` $\approx$ `r signif(liberalThreshold,1)` days before the most recent report date.
We assume that cases without a recorded death date (death date = NA) for more than `r signif(liberalThreshold,1)` after date of onset have recovered. This leaves `r signif(100*nrow(casesWhichWouldHaveDiedIfTheyWereGoingToDie)/nrow(transformedData),2)`\% of the cases to date which occurred early enough (date of onset at least `r signif(liberalThreshold,1)` before the last date of reporting) to be included in the case fatality rate.



Using more conservative requirement for inclusion, namely that delay between case onset and most recent report in the data set is greater than the maximum delay between onset and death (`r conservativeThreshold` days),  the case fatality is calculated to be `r caseFatalityEstimateUsingMaxTimeToDeath`\%.

The crude case fatality ($\frac{\text{n}_\text{deaths}}{\text{n}_\text{cases}}$) is `r crudeCaseFatality`\%.

```{r,echo=FALSE}
crudeCaseFatality = calculateCrudeCaseFatality(transformedData)
conservativeCaseFatality = calculateCarefulCaseFatality(transformedData, thresholdTimeToDeath = conservativeThreshold)

CFConfirmed = calculateCarefulCaseFatality(ebolaCasesConfirmed,thresholdTimeToDeath = liberalThreshold)
CFProbable = calculateCarefulCaseFatality(ebolaCasesProbable,thresholdTimeToDeath = liberalThreshold)
CFSuspected = calculateCarefulCaseFatality(ebolaCasesSuspected,thresholdTimeToDeath = liberalThreshold)
```

To be interpreted with some caution (due to smaller sample sizes): the case fatality rate per `status' is found to be (using the same method as described above):
Case fatality in 'Confirmed' cases: `r CFConfirmed`.
Case fatality in 'Probable' cases: `r CFProbable`.
Case fatality in 'Suspected' cases: `r CFSuspected`.

## Reporting delays
In most of the reported cases, there are delays between the date of onset and the date of report of the cases. The length of the delay may be a good indicator of the  effectiveness of the health system of the community. It could also be a useful metric when estimating the number of cases that might still be unreported. Figure 4 shows the distribution of the delay between date of onset and date of report. 

```{r, echo=FALSE}
hist(as.numeric(difftime(transformedData$reportDate,transformedData$onsetDate)),main="",xlab = "Reporting delay",ylab= "Frequency")
```

This SitRep should be interpreted in light of these reporting delays. For example, the incidence in the most recent week is likely under-reported to a significant extent - since $>$ half of cases take more than a `r signif(meanReportingDelay,1)` to be reported, a rough estimate would suggest that the incidence in the most recent week is approximately double the reported value.

## Time to death in fatal cases
Figure 5 presents the time delay from the onset of disease to the time of death in fatal cases.

```{r, echo=FALSE}
hist(as.numeric(transformedData$timeBeforeDeath),breaks = 10, main = '',xlab='Delays between onset an death in fatal cases',ylab= 'Frequency')
```

## Fitting an exponential growth model to the weekly incidence

An exponential growth model is fit to the weekly incidence data using the "fit" function from the R package "incidence". Weekly (rather than daily) incidence is used due to the sparsity of days with non-zero incidence during the early phase of the outbreak. These plots should be interpreted with caution due to the limited sample size of the available data.

A single exponential model (solid black lines) with confidence intervals (dashed lines) is plotted alongside the weekly incidence (red dots).


```{r, echo=FALSE}
weeklyIncidence = incidence(transformedData$onsetDate,interval=7)

model2 = fit(incidence(transformedData$onsetDate,interval=7),split = weeklyIncidence$dates[which.max(weeklyIncidence$counts)])

model = fit(incidence(transformedData$onsetDate,interval=7))

```

```{r,fig.height = 3, fig.width = 7, echo=FALSE}
plot(model) + geom_point(aes(x=incidence(transformedData$onsetDate,interval=7)$dates,y=incidence(transformedData$onsetDate,interval=7)$counts,col='red')) + theme(legend.position = "none")
```

The exponential daily growth rate in the single-model fit is estimated to be `r toString(signif(model$info$r,2))` with a 95\% confidence interval of (`r signif(model$info$r.conf[1],2)`; `r signif(model$info$r.conf[2],2)`). A positive growth rate value indicates that an epidemic is growing, whereas, a negative growth rate value shows that an epidemic is declining. 
We found that the single exponential model fit was flat, with a negative growth rate. This might suggest that the epidemic has peaked and  now getting smaller. 
To investigate this, we fitted two exponential functions to the incidence data. One for its exponential growth phase, and the other for the exponential decline phase of the epidemic. 
The overall pattern of the growth rate suggests that the epidemic grew slowly at the onset, and after reaching a peak, it started declining slowly almost in an identical pattern to its growth.   
The wide confidence bounds around the growth rate estimate is an indication that the small sample size may limit the accuracy of our estimates. Consequently,  these results should be interpreted with caution.

```{r,fig.height = 3, fig.width = 7, echo=FALSE}
plot(model2) + geom_point(aes(x=incidence(transformedData$onsetDate,interval=7)$dates,y=incidence(transformedData$onsetDate,interval=7)$counts,col='red')) + theme(legend.position = "none")
```

In the two-model fit, the exponential daily growth rates ($2.5\%$ - $95\%$ confidence intervals) leading up to the peak week and following the peak week are `r signif(model2$before$info$r,3)` (`r signif(model2$before$info$r.conf[1],3)`; `r signif(model2$before$info$r.conf[2],3)`) and `r signif(model2$after$info$r,3)` (`r signif(model2$after$info$r.conf[1],3)`; `r signif(model2$after$info$r.conf[2],3)`) respectively.



###Session info when this document was compiled
```{r,echo=FALSE}
sessionInfo()
```

###References



